Nested Stochastic Block Models Applied to the Analysis of Single Cell Data

### 2021-04-20 manuscript.tex
again optimal level

A possible strategy is based on Random Matrix Theory, as suggested by the authors of the SC3 package  \cite{kiselev_2017}, for which a suitable number of clusters, $\hat{k}$, is determined by the number of eigenvalues of the $\boldsymbol Z^\top\boldsymbol Z$ matrix (where $\boldsymbol Z$ is the normalized count matrix) significantly different at $p<.001$ from the appropriate Tracy-Widom distribution. According to this strategy, the optimal level $i$ is the one that minimizes the number of partitions and $\hat{k}$:

\begin{equation}
i = \underset{x}{\mathrm{argmin}} |B_x - \hat{k}|
\label{Equation_RMT}
\end{equation}

@@

An alternative strategy is to evaluate the behaviour of modularity at different hierarchy levels. While nSBM does not optimize the graph modularity, we observed that this tends to be maximal for the level better describing known cell populations.

@@

\begin{table}[h!]
\centering
 \begin{tabular}{|| l c c c c c ||}
%  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}
 \hline
 \textbf{Dataset} & \textbf{Cells} & \textbf{Edges} & \textbf{Leiden} & \textbf{PPBM} & \textbf{NSBM} \\ [0.5ex] 
 \hline\hline
 sc-mixology \cite{Tian_2019} & 860 & 9186 & 00:08 & 00:55 & 01:23 \\ 
 \hline
 Chromium 10x \cite{mereu_2020} & 1523 & 21447 & 00:24 & 01:34 & 03:29\\ 
 \hline
 Quartz-seq2 \cite{mereu_2020} & 1266 & 14603 & 00:16 & 00:30 & 02:10 \\
 \hline
 MARS-seq \cite{mereu_2020} & 1401 & 21756 & 00:24 & 01:10 & 06:11 \\
 \hline
 iCELL8 \cite{mereu_2020} & 1830 & 30636 & 00:34 & 01:31 & 09:06 \\
 \hline
 Paul15 \cite{paul_2015} & 2730 & 15444 & 00:34 & 02:53 & 06:55\\ 
 \hline
 Planaria \cite{plass_2018} & 21612 & 173667 & 06:32 & 34:12 & 3:17:08 \\
 \hline
% 4 & 545 & 18744 & 7560 \\
% \hline
% 5 & 88 & 788 & 6344 \\ [1ex] 
% \hline
\end{tabular}
\caption{Time required to minimise the nSBM using the default minimization method compared to the greedy merge-split MCMC. Times are expressed in mm:ss. Partition overlap measures concordance between the two models over the full hierarchy. Timing is the average after 3 initialisations.}
\label{Table1}
\end{table}

>>>>

A possible strategy is based on Random Matrix Theory, as suggested by the authors of the SC3 package  \cite{kiselev_2017}, for which a suitable number of clusters, $\hat{k}$, is determined by the number of eigenvalues of the $\boldsymbol Z^\top\boldsymbol Z$ matrix (where $\boldsymbol Z$ is the normalized count matrix) significantly different at $p<.001$ from the appropriate Tracy-Widom distribution. According to this strategy, the optimal level $i^k$ is the one that minimizes the number of partitions and $\hat{k}$:

\begin{equation}
i^k = \underset{x}{\mathrm{argmin}} |B_x - \hat{k}|
\label{Equation_RMT}
\end{equation}

@@

An alternative strategy is to evaluate the behaviour of modularity at different hierarchy levels. While nSBM does not optimize the graph modularity $Q$, we observed that this tends to be maximal for the level better describing known cell populations, so the optimal level $i^Q$ is

\begin{equation}
i^Q = \underset{x}{\mathrm{argmax}} |Q_x|
\label{Equation_MODmax}
\end{equation}

Where $Q_x$ is modularity at level $x$. We tabulated values arising from both the approaches in Table \ref{table_optimal}, together with the $i^D$ that is defined similarly to Eq. \ref{Equation_RMT} using the number of known annotated groups $D$ in place of $\hat{k}$

\begin{table}[h!]
\centering
 \begin{tabular}{|| l c c c c c c c c c ||}
 \hline
 \textbf{Dataset} & \textbf{Cells} & \textbf{D} & \textbf{D} & \textbf{nD} & \textbf{$\hat{k}$} & \textbf{$i$} & \textbf{$B_i$} & \textbf{Annotated groups} & \textbf{Annotated groups} \\ [0.5ex]

 \hline\hline
% sc-mixology \cite{Tian_2019} & 860 & 5 & 00:08 & 00:55 & 01:23 \\

\end{tabular}
\caption{foo bar}
\label{table_optimal}
\end{table}

@@

\begin{table}[h!]
\centering
 \begin{tabular}{|| l c c c c c ||}
%  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}
 \hline
 \textbf{Dataset} & \textbf{Cells} & \textbf{Edges} & \textbf{Leiden} & \textbf{PPBM} & \textbf{NSBM} \\ [0.5ex] 
 \hline\hline
 sc-mixology \cite{Tian_2019} & 860 & 9186 & 00:08 & 00:55 & 01:23 \\ 
 \hline
 Chromium 10x \cite{mereu_2020} & 1523 & 21447 & 00:24 & 01:34 & 03:29\\ 
 \hline
 Quartz-seq2 \cite{mereu_2020} & 1266 & 14603 & 00:16 & 00:30 & 02:10 \\
 \hline
 MARS-seq \cite{mereu_2020} & 1401 & 21756 & 00:24 & 01:10 & 06:11 \\
 \hline
 iCELL8 \cite{mereu_2020} & 1830 & 30636 & 00:34 & 01:31 & 09:06 \\
 \hline
 Paul15 \cite{paul_2015} & 2730 & 15444 & 00:34 & 02:53 & 06:55\\ 
 \hline
 Planaria \cite{plass_2018} & 21612 & 173667 & 06:32 & 34:12 & 3:17:08 \\
 \hline
% 4 & 545 & 18744 & 7560 \\
% \hline
% 5 & 88 & 788 & 6344 \\ [1ex] 
% \hline
\end{tabular}
\caption{Time required to run different partitioning strategies implemented in \texttt{schist} on various datasets. Times are expressed in mm:ss. }
\label{Table1}
\end{table}
